Case-based Arabic Morphological Analyzer for Arabic Text Mining

### 2010-12-17 trunk/tex/sarf.aclhlt.tex
reduce lines

Subfigures (a), (b), and (c) in Figure~\ref{f:example}
represent ${\cal P}$ the prefix ,
${\cal S}$ the stem, and ${\cal X}$ the suffix
transducers respectively. 
The symbol $\epsilon$ represents an empty string and is 
the source of non-determinism in this example.

@@

The case-based controller can interfere at any point in the 
analysis to make a decision like ignoring an FSA that 
is not interesting. 
It may also decide to correct the analysis of one FSA
based on the decision of the others. 
The controller uses the output of the transducers as input
to make its decision. 
It does not produce any external output and therefore 
the three transducers with the 
controller compose an FSA.

@@

Many affixes are composed of other affixes. For example,
the prefix \RL{wsy} is composed of three other prefixes
namely \RL{w}, \RL{s}, and \RL{y}.
The part of speech (POS) tag for \RL{wsy} is a concatenation
of the POS tags of each of the three morphemes. 
Keeping all valid morphemes in a list
is redundant and may lead to consistency issues when
users try to add their custom morphemes.
We introduce $L_{pp}$ and
$L_{xx}$ as prefix-prefix and suffix-suffix 
compatibility rules respectively.
This allows us to keep a shorter list of affixes. 
We also introduce a {\em resulting category}
for the $L_{pp}$ and  $L_{xx}$ resulting morphemes.
The resulting category in all practical cases
is that of the second morpheme in the pair.

Atomic affixes are standalone affixes
that are not the concatenation of other affixes. 
Formally, we define an affix $f$ to be either
an atomic affix $a$ or an atomic affix concatenated
with an affix. We write $f: a | a f$. 
%Sarf only keeps atomic affixes and their categories.

@@

We analyzed the prefixes and suffixes of 
Buckwalter~\shortcite{Buckwalter:02}
and constructed the recursive affixes
into directed acyclic graphs (DAG) similar to 
those in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
This transducer is deterministic and thus 
its traversal is linear.
This is computationally superior to the 
approach of Buckwalter where the analyzer considers
all possible substrings of the word in question
and looks it up in the affix tables.

@@

The algorithm \CodeIn{NDSarf} 
takes as input a text string $L$ and a case-based controller
$A$. 
It starts with one machine $\Phi_1$ similar to the one in 
Figure~\ref{f:composition} and inserts it into a collection
$M$. 
The algorithm reads a letter $c$ at a time from $L$
and iterates over all machines in $M$. 
For each $\Phi$ in $M$,
if $\Phi$ is in an accept state and it is in the suffix
phase, then the algorithm checks whether it should report
a full word. 
This happens when $c$ is a white space or a delimiter, 
or when the last letter leading to the current state
was a non-connecting letter.

In all cases, if $\Phi$ is in an accept state, 
it spawns another machine $\Psi$ and adds it to $M$. 
If $c$ leads to a reject state, then $\Phi$ dies 
and we remove it from $M$. 
Otherwise, $\Phi$ transitions using the $c$ edge.
After iterating over all machines in $M$, we invoke the 
controller $A$
and grant it full access and control over
all machines in $M$.

@@

We illustrate the hadith controller 
in Figure~\ref{f:hadith}. 
The transitions are labeled by outputs generated from 
the Sarf transducers. 
A NAME label denotes a proper name.
An NRC label denotes a narrator connector such as
\RL{`an} ``on behalf of'', \RL{.hada_t} ``narrated'', \RL{qaal} ``said'', 
\RL{'a_hbar} ``told'' or one of their derivations. 
The IBN label corresponds to the word \RL{ibn} ``the son of'' and is a name connector.
The NISBA label corresponds to an adjective that points to a person such 
as \RL{alma.sriy} ``the Egyptian''.

@@

We implemented our case study with Buckwalter in two phases. 
In phase one, we processed the entire text with Buckwalter and
obtained annotations to text. 
In the second phase, we fed the annotations to 
the hadith controller as if they were coming from Sarf. 
We measured the time it took to produce the annotations
and the time it took the controller to process the 
annotations without the time spent streaming the annotations.
We used the same setup for ElixirFM.

Sarf outperforms both Buckwalter and ElixirFM in terms of 
running time. 
It also produces higher accuracy measures. 
We think this is mainly due to the extended
lexicon as well as to the fact that Sarf reduces false positive
names by ignoring prefixes and suffixes that can not form
names even if one possible alternative of the stem is a name. 
Doing the same with Buckwalter and ElixirFM requires  
expanding the annotations resulting from one simple affix to be 
reflected in all the composite affixes that contain it.
%expanding the annotations resulting from one affix to reflect
%its multiple composing affixes.
%However, this was easily performed using our recursive definition 
%of the affixes.

Buckwalter performed much better than ElixirFM in 
terms of running time, 
while ElixirFM outperformed Buckwalter in terms of accuracy.
%This is up to 
%our knowledge the first comparison of these two tools and 
We think the big difference is due to the sophistication of 
grammar based functional morphological analysis ElixirFM 
uses and its reliance on additional annotations extracted from both ATB~\cite{Maamouri:04} and the Prague Arabic Dependency Treebank~\cite{Prague04}.

@@

We presented Sarf, a novel case-based morphological analyzer,
and validated its utility using the hadith literature
case study. 
Sarf uses non-deterministic finite state Automata with 
concatenative analysis and recursive affixes to analyze 
Arabic text. 
Our results showed that Sarf with the hadith controller
is faster and more accurate than
current morphological analyzers.

In the future, 
we will try Sarf with other interesting case studies.
We plan to complete the hadith case study 
and automate narrator discovery and authentication in the 
biography books. 
We plan to make use of the name connector and narrator
connector stop phrases that our case study extracted to create a 
learning system that can learn more names. 
We also plan on developing an Arabic linguistic computational model 
similar to that of ElixirFM~\cite{Otakar:07} but that can be 
explored partially based on the case study.

>>>>

Subfigures (a), (b), and (c) in Figure~\ref{f:example}
represent ${\cal P}$ the prefix ,
${\cal S}$ the stem, and ${\cal X}$ the suffix
transducers respectively. 
The symbol $\epsilon$ represents an empty string and is 
the source of non-determinism.

@@

The case-based controller can interfere at any point in the 
analysis to make a decision like ignoring an FSA that 
is not interesting. 
It may also decide to correct the analysis of one FSA
based on the decision of the others. 
The controller uses the output of the transducers as input
to make its decision. 
It does not produce external output and therefore 
the three transducers with the 
controller compose an FSA.

@@

Many affixes are composed of other affixes. For example,
the prefix \RL{wsy} is composed of three other prefixes
namely \RL{w}, \RL{s}, and \RL{y}.
The part of speech (POS) tag for \RL{wsy} is a concatenation
of the POS tags of each of the three morphemes. 
Keeping all valid morphemes in a list
is redundant and may lead to consistency issues when
users try to add their custom morphemes.
We introduce $L_{pp}$ and
$L_{xx}$ as prefix-prefix and suffix-suffix 
compatibility rules respectively.
This allows us to keep a shorter list of affixes. 
We also introduce a {\em resulting category}
for the $L_{pp}$ and  $L_{xx}$ resulting morphemes.
The resulting category in all practical cases
is that of the second morpheme in the pair. 
Atomic affixes are standalone affixes
that are not the concatenation of other affixes. 
Formally, we define an affix $f: a | a f$ to be either
an atomic affix $a$ or an atomic affix concatenated
with an affix. 
%Sarf only keeps atomic affixes and their categories.

@@

We analyzed the prefixes and suffixes of 
Buckwalter~\shortcite{Buckwalter:02}
and constructed the recursive affixes
into directed acyclic graphs (DAG) similar to 
those in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
This transducer is deterministic and thus 
its traversal is linear.
This is computationally superior to the 
approach of Buckwalter where the analyzer considers
all possible substrings %of the word in question
and looks them up in the affix tables.

@@

The algorithm \CodeIn{NDSarf} 
takes as input a text string $L$ and a case-based controller
$A$. 
It starts with one machine $\Phi_1$ similar to the one in 
Figure~\ref{f:composition} and inserts it into a collection
$M$. 
The algorithm reads a letter $c$ at a time from $L$
and iterates over all machines in $M$. 
For each $\Phi$ in $M$,
if $\Phi$ is in an accept state and it is in the suffix
phase, then the algorithm checks whether it should report
a full word. 
This happens when $c$ is a white space, a delimiter, 
or when the preceding letter %leading to the current state
was a non-connecting letter.

In all cases, if $\Phi$ is in an accept state, 
it spawns another machine $\Psi$ and adds it to $M$. 
Notice that spawning a new machine requires only keeping
the state of the machine and all machines share the
structures of the three transducers.
If $c$ leads to a reject state, then $\Phi$ dies 
and we remove it from $M$. 
Otherwise, $\Phi$ transitions using the $c$ edge.
After iterating over all machines in $M$, we invoke the 
controller $A$
and grant it full access and control over
all machines in $M$.

@@

We illustrate the hadith controller 
in Figure~\ref{f:hadith}. 
The transitions are labeled by outputs generated from 
the Sarf transducers. 
A NAME label denotes a proper name.
An NRC label denotes a narrator connector such as
\RL{`an} ``on behalf of'', \RL{.hada_t} ``narrated'', \RL{qaal} ``said'', 
\RL{'a_hbar} ``told'' or one of their derivations. 
An IBN label denotes the word \RL{ibn} ``the son of'' and is a name connector.
A NISBA label is a possessive adjective that qualifies a person such 
as \RL{alma.sriy} ``the Egyptian''.

@@

We implemented our case study with Buckwalter in two phases. 
In phase one, we processed the entire text with Buckwalter and
obtained annotations to text. 
In the second phase, we fed the annotations to 
the hadith controller as if they were coming from Sarf. 
We measured the time it took to accomplish both tasks
less the time spent streaming the annotations.
We did the same for ElixirFM.

Sarf outperforms both Buckwalter and ElixirFM in terms of 
running time. 
It also produces higher accuracy measures 
mainly due to the extended
lexicon as well as to the fact that Sarf reduces false positive
names by ignoring affixes that can not form
names even if one possible solution of the stem is a name. 
Doing the same with Buckwalter and ElixirFM requires  
expanding the annotations resulting from one simple affix to be 
reflected in all the composite affixes that contain it.
%expanding the annotations resulting from one affix to reflect
%its multiple composing affixes.
%However, this was easily performed using our recursive definition 
%of the affixes.

Buckwalter performed much better than ElixirFM in 
terms of running time, 
while ElixirFM outperformed Buckwalter in terms of accuracy.
%This is up to 
%our knowledge the first comparison of these two tools and 
We think the big difference is due to the sophistication of 
the formal functional morphological analysis of ElixirFM 
and its reliance on additional annotations extracted from 
both ATB~\cite{Maamouri:04} and the Prague Arabic Dependency Treebank~\cite{Prague04}.

@@

We presented Sarf, a novel case-based morphological analyzer,
and validated its utility using the hadith literature
case study. 
Sarf uses non-deterministic finite state Automata with 
concatenative analysis and recursive affixes to analyze 
Arabic text. 
Results show that Sarf with the hadith controller
is faster and more accurate than
current morphological analyzers.

In the future, 
we will try Sarf with other interesting case studies.
We plan to complete the hadith case study 
and automate narrator discovery and authentication in the 
biography books. 
%We plan to make use of the name connector and narrator
%connector stop phrases that our case study extracted to create a 
%learning system that can learn more names. 
%We also plan on developing an Arabic linguistic computational model 
%similar to that of ElixirFM~\cite{Otakar:07} but that can be 
%explored partially based on the case study.
